Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-08-03 improved_state_space.tex
Integrated Richard's revisions and shortened

\documentclass[12pt,english]{article}
\usepackage[verbose,letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} 
\linenumbers
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage{indentfirst}
\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
\usepackage[pdftex,hidelinks]{hyperref} 
\usepackage{rotating}
\usepackage[compact]{titlesec}
\titleformat{\section}[display]{\normalfont\bfseries}{}{0em}{\centering}
\titleformat{\subsection}[display]{\normalfont\itshape}{}{0em}{\centering}
\bibliographystyle{ecology}
\setlength{\bibsep}{0.0pt}
\bibpunct{(}{)}{,}{a}{}{,}
\setlength{\jot}{-1.2ex}
\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\doublespacing}
\makeatletter
\let\@fnsymbol\@arabic
\makeatother
\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data}

\author{Jeffrey A. Hostetler\thanks{Migratory Bird Center, Smithsonian 
Conservation Biology Institute, National Zoological Park, MRC 5503, 
Washington, DC 20013-7012} \textsuperscript{,}\footnote{Email: \href{mailto:hostetlerj@si.edu}{hostetlerj@si.edu}}
   \and Richard B. Chandler\footnote{USGS Patuxent Wildlife Research Center 
        Laurel, MD 20708} \textsuperscript{,}\thanks{Current address: Warnell School of
     Forestry and Natural Resources, University of Georgia}        
}
\date{} %remove date
\begin{document}
\maketitle
\vspace{-1cm}
\begin{spacing}{1.9}
\begin{flushleft}
\abstract{
Models of population dynamics 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by observation error, which
can inflate estimates of process variation if ignored.  
Recently, state-space models have been developed to deal with this problem
by directly modeling both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
several important limitations: (1) they assume that random effects are
Gaussian distributed, which implies that abundance can be negative and
that false positive observation errors are equally likely as false negative
errors; (2) they do not admit spatial variation in
population dynamics; and (3) some of the parameters of the model are not
estimable. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} 
that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects. 
We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. 
Results of a simulation study suggest that bias is 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
}

\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, 
%% open population point count models, 
population dynamics, 
%% random observer effects, 
Ricker-logistic model, zero-inflated

@@

\section*{Introduction}
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, 
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
In applied contexts, population models are used for estimating extinction risk % probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

Several challenges are routinely encountered when trying to fit models
of population dynamics to data from field studies. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

Recently developed state-space models 
have made it possible to
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a latent %hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be overestimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential. %% Exponential or geometric?
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \nonumber \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; %% Should this be \eta_t?
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian variables: %deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored, or
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

@@

With respect to the observation error, the Gaussian assumption
implies that false positive detections are
equally likely as false negatives, a pattern inconsistent with most
findings \citep{miller_etal:2011}. 
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as it being present but unavailable for detection,
or to individuals being available but missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised to account for %for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models of population dynamics 
%\citep[but see][]{buckland_etal:2004}. 
Of the exceptions that exist, none provide software to facilitate model fitting 
\citep{buckland_etal:2004}.

Other limitations of state-space models, as commonly applied in ecology,
include the failure to admit spatial variation in abundance and
the fact that some parameters are not identifiable
\citep{polansky_etal:2009}. 
Ignoring spatial variation reduces the scope of the
inferences and reduces the information about the parameters of interest \citep{dennis_etal:2010}.
%% Identifiability problems require 
%% model constraints or the assertion of additional assumptions. 
%% Specifically, 
Even more problematic is the issue that without some form of replication in the first year, 
abundance cannot be estimated without making strong assumptions about 
the initial distribution. 
%% because there is only one data
%% point (the count in the first year) that carries any information about
%%it. To remedy this, 
For example, practitioners either assume that there is no
observation error in the first year, or they assume the population is
at equilibrium, which may defeat the purpose of many studies of
population dynamics.

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. Observation models with
more intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of classical population growth models as well as state-space models.

@@

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance and detection parameters, 
%% associated with both abundance and the detection process, 
the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

Regardless of the specified distribution, the model for initial
abundance has two distinguishing features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
\beta^{\Lambda}_1
x_{i}$. Second, the spatial
replicates resolve the
problem of parameter non-identifiability that are common to
standard state-space models because
%, as demonstrated by \citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data \citep{royle:2004biom}. Hence, the first component of
the model addresses both the issues of spatial inference and
parameter identifiability discussed previously.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$
%% , which simulates 
of geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
\begin{samepage}
Partitioning population growth into survival and recruitment
is useful %%ideal 
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: %% Should this geometric instead of exponential?
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker-logistic population growth model %% Could we just call this the "Ricker model"?
\citep{ricker:1954}. Another option is a modified Gompertz-logistic
density-dependent model \citep{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker-logistic model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.
\end{samepage}

@@

One limitation of the geometric-recruitment, exponential,
Ricker-logistic, and Gompertz-logistic versions of the DM models %%all
is that they include no mechanism for a local population to recover
after extinction. 
However, these models can be generalized 
to include both internal and external (immigration) contributions
to population growth. For example, an exponential plus immigration
model is:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
\end{linenomath*}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.

@@

The negative binomial distribution used in $N$-mixture models can have
a very long right tail, and its mean to variance ratio may not be
ideal in all cases. An alternative is to consider zero-inflated
distributions, such as the zero-inflated Poisson:
\begin{linenomath*}
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
\mbox{Pois}(0) &\; \text{with probability} \; \psi \\
\mbox{Pois}(\Lambda) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
\end{linenomath*}
where $\psi$ represents the proportion of extra zeros. 
%% Mechanistically, this distribution might be appropriate when not all sites are 
%% within the range of a species.
%%
Under this formulation, 
detection, abundance, and zero-inflation can be modeled separately as
functions of covariates. For example, detection of a
species might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.

@@

\subsection*{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, 
it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\mathrm{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
% NOTE: I would save this for the discussion
% Conceptually, this is an appropriate method of modeling demographic stochasticity 
% for semelparous organisms with one generation per sampling interval (year)
% \citep{bonsall_hastings:2004,melbourne_hastings:2008},
% but may actually perform better for iteroparous organisms as well (see Discussion).  
% Dynamics of semelparous organisms with more than one generation per year 
% (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
% year is known, then this can be accounted for in our models by adding extra $N$'s between each
% count (see Supplement 1).

>>>>

\documentclass[12pt,english]{article}
\usepackage[verbose,letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} 
\linenumbers
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage{indentfirst}
\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
\usepackage[pdftex,hidelinks]{hyperref} 
\usepackage{rotating}
\usepackage[compact]{titlesec}
\titleformat{\section}[display]{\normalfont\bfseries}{}{0em}{\centering}
\titleformat{\subsection}[display]{\normalfont\itshape}{}{0em}{\centering}
\bibliographystyle{ecology}
\setlength{\bibsep}{0.0pt}
\bibpunct{(}{)}{,}{a}{}{,}
\setlength{\jot}{-1.2ex}
\setstretch{1}
\widowpenalty 2
\clubpenalty 2
\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\doublespacing}
\makeatletter
\let\@fnsymbol\@arabic
\makeatother
\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data}

\author{Jeffrey A. Hostetler\thanks{Migratory Bird Center, Smithsonian 
Conservation Biology Institute, National Zoological Park, MRC 5503, 
Washington, DC 20013-7012} \textsuperscript{,}\footnote{Email: \href{mailto:hostetlerj@si.edu}{hostetlerj@si.edu}}
   \and Richard B. Chandler\footnote{USGS Patuxent Wildlife Research Center 
        Laurel, MD 20708} \textsuperscript{,}\thanks{Current address: Warnell School of
     Forestry and Natural Resources, University of Georgia}        
}
\date{} 
\begin{document}
\maketitle
\vspace{-1cm}
\begin{spacing}{1.9}
\begin{flushleft}
\abstract{
Models of population dynamics 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by observation error, which
can inflate estimates of process variation if ignored.  
Recently, state-space models have been developed to deal with this problem
by directly modeling both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
several important limitations: (1) they assume that random effects are
Gaussian distributed, which implies that abundance can be negative and
that false positive observation errors are equally likely as false negative
errors; (2) they do not admit spatial variation in
population dynamics; and (3) some of the parameters of the model are not
estimable. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} 
that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects. 
We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. 
Results of a simulation study suggest that bias is 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
}

\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic, immigration, % model
population dynamics, Ricker-logistic, zero-inflated % model

@@

\section*{Introduction}
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, 
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
In applied contexts, population models are used for estimating extinction risk 
\citep{nadeem_lele:2011,hostetler_etal:2012} and for predicting the %schoener_spiller:1992,
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

Two challenges are routinely encountered when trying to fit models %Several
of population dynamics to data from field studies. 
First, deterministic models of population dynamics are generally inadequate due to process %virtually always
variation, the stochasticity in demographic parameters and environmental % inherent
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

Recently developed state-space models 
have made it possible to
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a latent 
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be overestimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model (henceforth Ricker model), or it might be 
density-independent, such as when growth
is exponential. %% Exponential or geometric?
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= x_0 \nonumber \\ %X_1
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
where $x_0$ is the model for initial abundance.  In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian variables: 
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored, or
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

@@

With respect to the observation error, the Gaussian assumption
implies that false positive detections are
equally likely as false negatives, a pattern inconsistent with most
findings \citep{miller_etal:2011}. 
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as it being present but unavailable for detection,
or to individuals being available but missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised to account for 
this form of observation error, rarely have these methods been
integrated into state-space models of population dynamics. 
Of the exceptions that exist, none provide software to facilitate model fitting 
\citep{buckland_etal:2004}.

Other limitations of state-space models, as commonly applied in ecology,
include the failure to admit spatial variation in abundance and
the fact that some parameters are not identifiable
\citep{polansky_etal:2009}. 
Ignoring spatial variation reduces the scope of the
inferences, reduces the information about the parameters of interest,
and can lead to underestimation of variances \citep{dennis_etal:2010}.
Even more problematic is the issue that without some form of replication in the first year, 
abundance cannot be estimated without making strong assumptions about 
the initial distribution. 
For example, practitioners either assume that there is no
observation error in the first year ($x_0 = X_1$), or they assume the population is
at equilibrium, which may defeat the purpose of many studies of
population dynamics.

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. An observation model %with
%more intuitive interpretations, such as those that 
explicitly modeling
detection probability has been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of classical population growth models as well as state-space models.

@@

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance and detection parameters, 
the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

Regardless of the specified distribution, the model for initial
abundance has two distinguishing features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
\beta^{\Lambda}_1
x_{i}$. Second, the spatial
replicates resolve the
problem of parameter non-identifiability that are common to
standard state-space models because
process variation and observation error can be estimated from
spatially-replicated count data \citep{royle:2004biom}. Hence, the first component of
the model addresses both the issues of spatial inference and
parameter identifiability discussed previously.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new %
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma N_{i,t-1})$
of geometric or density independent growth; and the
no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be accommodated, e.g., % easily %for example
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
%\begin{samepage}
Partitioning population growth into survival and recruitment
is useful  
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: %% Should this geometric instead of exponential?
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker model \citep{ricker:1954}. Another option is a 
modified Gompertz-logistic
density-dependent model \citep[henceforth Gompertz model,][]{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.
%\end{samepage}

@@

One limitation of the geometric-recruitment, exponential,
Ricker, and Gompertz versions of the DM models 
is that they include no mechanism for a local population to recover
after extinction. 
However, these models can be generalized 
to include both internal and external (immigration) contributions
to population growth. For example, an exponential plus immigration
model is:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
  \label{eq:expimm2}
\end{equation}
\end{linenomath*}
where $\iota$ represents the average number of immigrants per year, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  
The geometric-recruitment, Ricker,
and Gompertz models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker + immigration and the 
Gompertz + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.

@@

The negative binomial distribution used in $N$-mixture models can have
a very long right tail, and its mean to variance ratio may not be
ideal in all cases. An alternative is to consider zero-inflated
distributions, such as the zero-inflated Poisson:
\begin{linenomath*}
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
\mbox{Pois}(0) &\; \text{with probability} \; \psi \\
\mbox{Pois}(\Lambda) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
\end{linenomath*}
where $\psi$ represents the proportion of extra zeros. 
Under this formulation, 
detection, abundance, and zero-inflation can be modeled separately as
functions of covariates. For example, detection of a
species might depend on wind speed,
abundance on forest type and weather, and zero-inflation upon elevation and
climate.

@@

\subsection*{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, 
it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\mathrm{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
% NOTE: I would save this for the discussion
Conceptually, this is an appropriate method of modeling demographic stochasticity 
 for semelparous organisms with one generation per sampling interval (year)
 \citep{bonsall_hastings:2004},
and the DM models appropriate for modeling the demographic stochasticity of iteroparous organisms,
if they are underdispersed compared to the Poisson (but see Discussion).
% but may actually perform better for iteroparous organisms as well (see Discussion).  
 Dynamics of semelparous organisms with more than one generation per year 
 (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
 year is known, then this can be accounted for in our models by adding extra $N$'s between each
 count (see Supplement 1).

One way to model environmental stochasticity --- variation in population growth
rate due to stochastic environmental conditions --- is by explicitly
modeling dynamic parameters as functions of environmental covariates. %, such as weather.  
For example, in the Gompertz model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature  
and $K$ a function of rainfall.

Another way to model environmental stochasticity is as lognormal variation in 
the population growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  
This could be applied to both density-independent and density-dependent models.
For example, for the Ricker + immigration model:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
\end{linenomath*}
where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
We suggest three variations for spatial variation in environmental stochasticity.  
The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second, 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
for $\nu_{i,t}$ with distance-dependent correlation coefficients 
between sites in a year might be appropriate.
